---
title: "simulating serosurveys"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulating_serosurveys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(serofoi)
library(ggplot2)
library(dplyr)
library(purrr)
library(forcats)
```

In this vignette, we show how it is possible to simulate serosurveys using the `simulate_serosurvey` function. This function separates two aspects: the serocatalytic model used to simulate population-wide seropositivity throughout individuals' lives; and the features of the particular serological survey that are being used to uncover these dynamics.

## Constant FOI
We start by assuming that a disease has a constant force-of-infection (FOI) over time.

```{r}
max_age <- 80
foi_df <- data.frame(
  age = seq(1, max_age, 1),
  foi = rep(0.05, max_age)
)

foi_df %>% 
  ggplot(aes(x=age, y=foi)) +
  geom_line()
```
We suppose that we carry out a serological survey which randomly samples the population. For simplicity, to start, we imagine that all individual one-year group ages are sampled and the sample size in each age is the same: $n=15$.

```{r}
sample_size <- 15
survey_features <- data.frame(
  age_min = seq(1, max_age, 1),
  age_max = seq(1, max_age, 1),
  sample_size = rep(sample_size, max_age))
```

We then put these pieces together and simulate a serosurvey. Either a time-varying FOI model or an age-varying FOI model would produce the same results here since they are equivalent when the FOI is constant.
```{r}
serosurvey <- simulate_serosurvey(
  model = "age",
  foi = foi_df,
  survey_features = survey_features
)

glimpse(serosurvey)
```
And we can then plot the proportions of those seropositive. 
```{r}
serosurvey %>% 
  ggplot(aes(x=age_min, y=n_seropositive)) +
  geom_point() +
  geom_smooth(se=FALSE)
```

We can also simulate the true seropositivities and overlay these on the sample-based equivalents.
```{r}
seropositive_true <- probability_seropositive_by_age(
  model = "age",
  foi = foi_df,
  seroreversion_rate = 0
) %>% 
  mutate(age_min = age)

serosurvey %>% 
  left_join(seropositive_true, by='age_min') %>% 
  mutate(seropositivity_sampled=n_seropositive/sample_size) %>% 
  ggplot(aes(x=age_min, y=seropositivity_sampled)) +
  geom_point() +
  geom_line(aes(y=seropositivity),
            colour="orange") +
  ylab("Seropositivity")
```

## Age-varying FOI
We now consider a pathogen which has an FOI that varies throughout the age of individuals. We assume the FOI declines strongly throughout childhood.

```{r}
foi_df <- data.frame(
  age = seq(1, max_age, 1)
) %>% 
  mutate(foi=0.1 * exp(-0.1 * age))

foi_df %>% 
  ggplot(aes(x=age, y=foi)) +
  geom_line()
```
We use the same survey design as before and simulate a serological survey.

```{r}
serosurvey_age_dep <- simulate_serosurvey(
  model = "age",
  foi = foi_df,
  survey_features = survey_features
)

# combine with constant FOI survey
serosurvey_combined <- serosurvey %>% 
  mutate(type="constant FOI") %>% 
  bind_rows(serosurvey_age_dep %>% 
              mutate(type = "age-dependent FOI")) %>% 
  mutate(type=as.factor(type)) %>% 
  mutate(type=fct_relevel(type,
                          "constant FOI",
                          "age-dependent FOI")) 

# plot both
serosurvey_combined %>% 
  ggplot(aes(x=age_min, y=n_seropositive)) +
  geom_point() +
  geom_smooth(se=FALSE) +
  facet_wrap(~type)
```
Above, we see that the FOI for the age-dependent FOI pathogen increases rapidly during the earliest years then plateaus in later adulthood as the FOI effectively becomes negligible.

## Time-dependent FOI
We now consider a pathogen where the FOI varies only over calendar time. We imagine a pathogen that has a 'spiky' FOI, characteristic of an epidemics.

```{r}
spiky_foi <- c(
  rep(0, 10),
  rep(0.1, 2),
  rep(0, 20),
  rep(0.2, 1),
  rep(0, 15),
  rep(0.1, 2),
  rep(0, 30)
)

foi_df <- data.frame(
  year = seq(1946, 2025, 1),
  foi = spiky_foi
)

# plot
foi_df %>% 
  ggplot(aes(x=year, y=foi)) +
  geom_line()
```
The same serological survey design then produces a serological profile with distinct patterning.
```{r}
serosurvey <- simulate_serosurvey(
  model = "time",
  foi = foi_df,
  survey_features = survey_features
)

# plot shows jumps in seropositivity
serosurvey %>% 
  ggplot(aes(x=age_min, y=n_seropositive)) +
  geom_point() +
  geom_smooth(se=FALSE)
```
## Age-and-time-dependent FOI
Many pathogens may have a transmission strength that varies according to both age and time. A good example of this could be for a sexually transmitted disease where transmission follows a characteristic age-specific pattern, peaking in the early 20s. Here, we imagine such a disease which has only recently invaded a population.

We imagine the time-specific multiplier of FOI follows the below variation.
```{r}
foi_time <- c(rep(0, 40), rep(1, 40))
foi_df_time <- data.frame(
  year = seq(1946, 2025, 1),
  foi = foi_time
)

# plot
foi_df_time %>% 
  ggplot(aes(x=year, y=foi)) +
  geom_line()
```

We create a pattern of age-structured FOI multipliers which peaks in those of early 20s.
```{r}
ages <- seq(1, 80, 1)
foi_age <- 2 * dlnorm(
  ages, meanlog = 3.5, sdlog = 0.5)

foi_df_age <- data.frame(
  age = ages,
  foi = foi_age
)

# plot
foi_df_age %>% 
  ggplot(aes(x=age, y=foi)) +
  geom_line()
```

To create the overall FOI (which is a function of both time and age), we create the product of the time- and age-dependent parts of it.
```{r}
foi_df <- expand.grid(
  year=foi_df_time$year,
  age=foi_df_age$age
) %>% 
  left_join(foi_df_age, by="age") %>% 
  rename(foi_age=foi) %>% 
  left_join(foi_df_time, by="year") %>% 
  rename(foi_time=foi) %>% 
  mutate(foi = foi_age * foi_time) %>% 
  select(-c("foi_age", "foi_time")) %>% 
  mutate(birth_year = year - age) %>% 
  filter(birth_year >= 1945) %>% 
  arrange(birth_year, age)

# plot a few birth years
select_birth_years <- c(1955, 1965, 1975, 1985, 1995, 2005, 2015)
foi_df %>% 
  filter(birth_year %in% select_birth_years) %>% 
  mutate(birth_year = as.factor(birth_year)) %>% 
  ggplot(aes(x=year, y=foi, colour=birth_year)) +
  geom_line()
```

We now simulate a serosurvey assuming these historical FOIs generated the population-wide seropositivities in 2025; we make the sample sizes larger to be able to clearly visualise the patterns in seropositivity.
```{r}
max_age <- 80
sample_size <- 50
survey_features <- data.frame(
  age_min = seq(1, max_age, 1),
  age_max = seq(1, max_age, 1),
  sample_size = rep(sample_size, max_age))

serosurvey <- simulate_serosurvey(
  model = "age-time",
  foi = foi_df %>% select(-birth_year),
  survey_features = survey_features
)

# plot
serosurvey %>% 
  ggplot(aes(x=age_min, y=n_seropositive)) +
  geom_point() +
  geom_smooth(se=FALSE)
```

