---
title: "An Introduction To Force-of-Infection Models"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r cleaning, include = FALSE, echo = TRUE}
library(serofoi)
library(dplyr)
```

The current version of ***serofoi*** supports three different types of serocatalytic models for estimating the *Force-of-Infection* (FoI): constant, age-varying, and time-varying FoI. To estimate relevant parameters we employ a suit of Bayesian models using age-disaggregated population-based cross-sectional serological surveys as input data.

## What is the FoI

The FoI, also known as the hazard rate or the infection pressure, is a key concept in mathematical modelling of infectious diseases. It represents the rate at which susceptible individuals become infected given their exposure to a pathogen. In simple terms, the FoI quantifies the risk of a susceptible individual becoming infected over a period of time. It is usually expressed as a rate per unit of time (e.g., per day or per year).

To illustrate this, consider a simple SIR model:
$$
\begin{aligned}
    \frac{dS}{dt} &= -\beta S I,\\
    \frac{dI}{dt} &= \beta S I - \gamma I,\\
    \frac{dR}{dt} &= \gamma I,
\end{aligned}
$$

In this case, the FoI is defined as $\lambda(t) = \beta I(t)$ (Note that the FoI is time dependent by definition). To solve this system of equations we need to know the transmission rate $\beta>0$, and the rate of recovery from infection $\gamma>0$, as well as some initial conditions.

**Serocatalytic models** offer a simpler perspective. Here we are interested in estimating how the risk of historical exposure to a pathogen varies depending on the time of birth of the individuals in the sample. For this, we put together the $I$ and $R$ compartments into a single **seropositives** compartment $P(t)$ and consider that the dynamics of each birth cohort separately

$$
\begin{aligned}
    &\frac{dN^{\tau}(t)}{dt} = -\lambda(t) N^{\tau}(t),\\
    &\frac{dP^{\tau}(t)}{dt} = \lambda(t) N^{\tau}(t), \\
    &N^{\tau}(t)+P^{\tau}(t)=1; \quad t<\tau
\end{aligned}
$$
where $0 \leq N^\tau(t)\leq 1$ is the proportion of individuals born in year $\tau$ who are seronegative at time $t$. Assuming lack of maternal antibodies, the initial condition for each cohort can be specified as $N^{\tau}(\tau) = 1$. Depending on whether the FoI is constant, time dependent, or age dependent, this system of equations yields the different solutions on which *serofoi* is based on.

## Constant vs Time-varying FoI

The FoI is often incorrectly assumed to be constant over time. Identifying whether the FoI follows a constant, an age-varying or a time-varying trend can be important in the identification and characterization of the spread of disease. **serofoi** offers tools to implement and compare a wide variety of Bayesian models to estimate the FoI according to the aforementioned serocatalytic models.

In the idealised situation where the FoI is constant $\lambda(t) = \lambda$:

$$
\frac{dP^{\tau}(t)}{dt} = \lambda (1 - P^{\tau}(t)).
$$
Applying the integrating factor method and using the initial conditions $X^\tau(\tau) = 0$ yields to the solution:

$$
P^{\tau}(t) = 1 - \exp(-\bar \lambda t),
$$

meaning that, in the long run, the seropositivity converges to 1.

In more realistic applications, the FoI is assumed to be piecewise-constant on discrete intervals of 1-year length. The solution for each year is then:
$$
P^{\tau}(t) = 1 + (P_0 - 1) \exp(-\lambda_t t)
$$
where $P_0$ corresponds to the seropositivity at the end of the previous year and $\lambda_t$ is the constant value assumed for the FoI at time $t$. From the resulting recursive equation, it is possible to obtain analytical solutions for the seropositivity $P^\tau(t)$ along each 1-year length chunk. 

## Considering seroreversion

Since it is possible for seropositive individuals to lose immunity over time, we consider the rate at which infected individuals become seronegative $\mu$. The system of equations is now:

$$
\begin{aligned}
    \frac{dP^{\tau}(t)}{dt} &= \lambda(t) (1 - P^{\tau}(t)) - \mu P^{\tau}(t),\\
    P^{\tau}(\tau) &= 0,
\end{aligned}
$$

Similarly, this yields to a iterative solution for the seropositivity $P^\tau (t)$ when piece-wise constant FoI is considered:


$$
  P^{\tau}(t)  = \frac{\lambda_t}{\lambda_t + \mu} + \left(P_0 - \frac{\lambda_t}{\lambda_t + \mu}\right) \exp(-(\lambda_t + \mu)),
$$

where, again, $P_0$ corresponds to the proportion of seropositive individuals of cohort $\tau$ by the end of the previous year. Correct indexation and simplification of this equation allows the description of age-dependent FoIs.

## Estimating the FoI - Bayesian modelling

Now that we can describe the proportion of seropositive individuals $P(t)$ by means of the serocatalytic models for constant, time-varying, and age-varying FoIs, what we want is to obtain estimates of the FoI ($\lambda$) and the seroreversion rate ($\mu$) by sampling bye means of Bayesian methods taking a cross-sectional serological survey as data. For this, we can use **serofoi**'s `fit_seromodel()`, which relies on the statistical programming language *Stan* to perform this task. The current available options for modelling are summarised in table 1:


::: l-body-outset
| `model_type` | scale (`is_log_foi`)  |  $\lambda_1$   | $\lambda_i$, $i>1$  | $\mu$     | $\sigma$ |
| ------------ | --------------------- | -------------- | ------------------- | --------- | -------- |
| `"constant"` | `FALSE`               | $\sim\mathcal{N}(\bar{\lambda},\sigma)$ or <br>$\sim\mathcal{U(\lambda_{min},\lambda_{max})}$ | \-| $\sim\mathcal{N}(\bar{\mu},\sigma_\mu)$ or <br>$\sim\mathcal{U(\mu_{min},\mu_{max})}$ | $\mathcal{C}(\sigma_0, \gamma)$ |
| `"age"`      | `FALSE`               | $\sim\mathcal{N}(\bar{\lambda}_1,\sigma)$ or <br> $\sim\mathcal{U(\lambda_{1,min},\lambda_{1,max})}$ | $\sim\mathcal{N}(\lambda_{i-1},\sigma)$| $\sim\mathcal{N}(\bar{\mu},\sigma_\mu)$ or <br> $\sim\mathcal{U(\mu_{min},\mu_{max})}$ | $\mathcal{C}(\sigma_0, \gamma)$ |
| `"time"`     | `FALSE`               | $\sim\mathcal{N}(\bar{\lambda}_1,\sigma)$ or <br> $\sim\mathcal{U(\lambda_{1,min},\lambda_{1,max})}$ | $\sim\mathcal{N}(\lambda_{i-1},\sigma)$ |$\sim\mathcal{N}(\bar{\mu},\sigma_\mu)$ or <br> $\sim\mathcal{U(\mu_{min},\mu_{max})}$ | $\mathcal{C}(\sigma_0, \gamma)$ |
| `"time"`                | `TRUE`     | $\sim\mathcal{N}(\bar{\lambda}_1,\sigma)$ or <br> $\sim\mathcal{U(\lambda_{1,min},\lambda_{1,max})}$ | $\sim\mathcal{N}(\lambda_{i-1},\sigma)$ | $\sim\mathcal{N}(\bar{\mu},\sigma_\mu)$ or <br> $\sim\mathcal{U(\mu_{min},\mu_{max})}$ | $\mathcal{C}(\sigma_0, \gamma)$ |
:::
Table 1. Available options. Here, $\mathcal{N}$ corresponds to the normal distribution, $\mathcal{U}$ to the uniform distribution, and $\mathcal{C}$ to the Cauchy distribution.

The number of seropositives at the time of the survey $X(a)$, corresponding to `n_seropositive` in the data, follows a binomial distribution:

$$
  X(a) \sim Binom\left(n(a), P^{T - a} (T)\right)
$$

where $n(a)$ corresponds to the sample size for age group $a$ (`n_sample`) and $T$ is the time when the serosurvey was conducted. Note that $T-a$ correspond to the birth year of the cohort with age $a$ at the time of the survey.

The prior distributions for each parameter can be selected in `fit_seromodel()`. This is done by means of the helper functions `sf_normal()`, ``sf_uniform()` and `sf_cauchy()`. For example:

```{r prior_example, eval=FALSE}
fit_seromodel(
  serosurvey,
  model_type = "time",
  foi_prior = sf_uniform(0.5, 1),
  foi_sigma_rw = sf_cauchy(0.01, 0.005),
  is_seroreversion = TRUE,
  seroreversion_prior = sf_normal(0.02, 0.01)
)
```

Corresponds to the model:
$$
\begin{align}
  \lambda_1 &\sim \mathcal{U}(0.5, 1) \\
  \sigma &\sim \mathcal{C}(0.01, 0.005) \\
  \mu &\sim \mathcal{N}(0.02, 0.01)
\end{align}
$$
The available options for each parameter are shown in table 1. Note that, in the case of the age- and time-varying models, the FoI prior that can be specified is that of the first FoI value to be estimated (i.e. the birth year of the oldest age-group in the time-varying case and the youngest age-group in the age-varying case), such that the means for subsequent ages/years are selected iteratively as the FoI value estimated for the previous index.

## Model 1. Constant Force-of-Infection (endemic model)
The *endemic constant model* is a simple mathematical model used in epidemiology to describe the seroprevalence of an infectious disease within a population, as a product of a long-term transmission. In this case, the rate of infection acquisition $\lambda$ is constant over time, and the seroprevalence $P(A)$ behaves as a cumulative process increasing monotonically with age.

To test the `constant` model we simulate a serosurvey following a stepwise decreasing FoI (red line in Fig. 1) using the data simulation functions of ***serofoi***:

```{r constant simdata, include = TRUE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE}
# how the disease circulates
foi_df <- data.frame(
    year = seq(2000, 2049, 1),
    foi = rep(0.02, 50)
)

# specify 5-year age bins for observations
survey_features <- data.frame(
    age_min = seq(1, 50, 5),
    age_max = seq(5, 50, 5),
    n_sample = rep(25, 10)
)

serosurvey_constant <- simulate_serosurvey(
  "time",
  foi_df,
  survey_features
) |>
mutate(survey_year = 2050)
```

The simulated dataset `serosurvey_constant` contains information about 250 samples of individuals between 1 and 50 years old (5 samples per age) with age groups of 5 years length. The following code shows how to implement the constant model to this simulated serosurvey:

```{r constant model, fig.cap = "Figure 1. Constant serofoi model plot. Simulated (red) vs modelled (blue) FoI.", include = TRUE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE, fig.width=4, fig.asp=1.5, fig.align="center", out.width ="50%", fig.keep="all"}
seromodel_constant <- fit_seromodel(
  serosurvey = serosurvey_constant,
  model_type = "constant",
  iter = 800
)
plot_seromodel(
  seromodel_constant,
  serosurvey = serosurvey_constant,
  foi_df = foi_df,
  size_text = 6
)
```

```{r clean env constant, include = FALSE, echo = TRUE, message=FALSE}
rm(list = ls(pattern = "_constant"))
```

In this case, 800 iterations are enough to ensure convergence. `plot_seromodel()` returns a visualisation of the results including a summary where the expected log pointwise predictive density (`elpd`) and its standard error (`se`) are shown. We say that a model converges if all the R-hat estimates are below 1.01.

## Time-varying FoI models

For the time-varying FoI models, the probability for a case to be positive at age a at time $t$ also follows a binomial distribution, as described above. However, the seroprevalence is obtained from a cumulative of the yearly-varying values of the FoI over time:

$$
  P^{\tau}(t)  = \frac{\lambda_t}{\lambda_t + \mu} + \left(P_0 - \frac{\lambda_t}{\lambda_t + \mu}\right) \exp(-(\lambda_t + \mu)),
$$

## Model 2. Time-varying FoI - Slow Time-Varying FoI

The *time-varying model* (`model_type = "time"`) uses a forward random walk algorithm where the prior of the first chronological FoI value in the time-span of the serological survey can be either a normal distribution (`foi_prior = sf_normal()`) or a uniform distribution (`foi_prior = sf_uniform()`). Subsequent values of the FoI are sampled from the normal distribution as $\lambda_i \sim \mathcal{N}(\lambda_{i-1}, \sigma)$, where $\sigma \sim \mathcal{C}(0, 1)$. The prior distribution for the seroreversion rate can also be either normal or uniform, and is specified by setting `is_seroreversion = TRUE` and by means of the parameter `seroreversion_prior`.

To test the `"time"` using a normal prior in the FoI, we simulate a serosurvey following a stepwise decreasing FoI (red line in Fig. 2) using the data simulation functions of ***serofoi***:

```{r tv_normal simdata, include = TRUE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE}

foi_df <- data.frame(
    year = seq(2000, 2049, 1),
    foi = c(
      rep(0.2, 25),
      rep(0.1, 10),
      rep(0.00001, 15)
    )
)

survey_features <- data.frame(
    age_min = seq(1, 50, 5),
    age_max = seq(5, 50, 5),
    n_sample = rep(25, 10)
)

serosurvey_sw_dec <- simulate_serosurvey(
  "time",
  foi_df,
  survey_features
) |>
mutate(survey_year = 2050)

```

The simulated  dataset `serosurvey_sw_dec` contains information 250 samples of individuals between 1 and 50 years old (5 samples per age) with age groups of 5 years length. The following code shows how to implement the slow time-varying normal model to this simulated serological survey:

```{r tv_normal model, fig.cap = "Figure 2. Slow time-varying serofoi model plot.  Simulated (red) vs modelled (blue) FoI.", include = TRUE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE, fig.width=4, fig.asp=1.5, fig.align="center", out.width ="50%", fig.keep="all"}
foi_index <- data.frame(
  year = seq(2000, 2049),
  foi_index = rep(c(1, 2, 3), c(25, 10, 15))
)
seromodel_time_normal <- fit_seromodel(
  serosurvey = serosurvey_sw_dec,
  model_type = "time",
  foi_index = foi_index,
  iter = 1500
)
plot_seromodel(
  seromodel_time_normal,
  serosurvey = serosurvey_sw_dec,
  foi_df = foi_df,
  size_text = 6
)
```

```{r clean env sw_dec, include = FALSE, echo = TRUE, message=FALSE}
rm(list = ls(pattern = "_sw_dec|_normal"))
```

The number of iterations required may depend on the number of years, reflected by the difference between the year of the serosurvey and the maximum age-class sampled.

## Model 3. Time-varying FoI - Fast Epidemic Model
The *time-varying fast epidemic model* is parametrized in such a way that the initial FoI value in the forward random walk is sampled in the logarithmic scale. In this case, the priors for subsequent FoIs are:

$$
\begin{aligned}
  \log\left(\lambda_t\right) &\sim \mathcal{N}(\log(\lambda_{t-1}), \sigma) \\
  \sigma &\sim \mathcal{N} (0, 1)
\end{aligned}
$$

This is done in order to capture fast changes in the FoI trend. 

To test the `log_time` model we simulate a serosurvey conducted in 2050 emulating a hypothetical situation where a three-year epidemic occurred between 2030 and 2035:

```{r tv_normal_log simdata, include = TRUE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE}
foi_df <- data.frame(
    year = seq(2000, 2049, 1),
    foi = c(
      rep(0, 30),
      rep(0.7, 3),
      rep(0, 17)
    )
)

survey_features <- data.frame(
    age_min = seq(1, 50, 5),
    age_max = seq(5, 50, 5),
    n_sample = rep(25, 10)
)

serosurvey_large_epi <- simulate_serosurvey(
  survey_features = survey_features,
  foi_df,
  model = "time"
) |>
mutate(survey_year = 2050)

```

The simulated serosurvey tests 250 individuals between 1 and 50 years old by the year 2050. The implementation of the fast epidemic model can be obtained running the following:

```{r tv_normal_log model, fig.cap = "Figure 3. *Time-varying fast epidemic serofoi model* plot. Simulated (red) vs modelled (blue) FoI.", include = TRUE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE, fig.width=4, fig.asp=1.5, fig.align="center", out.width ="50%", fig.keep="all"}
foi_index <- data.frame(
  year = seq(2000, 2049),
  foi_index = rep(c(1, 2, 3), c(30, 3, 17))
)
seromodel_log_time_normal <- fit_seromodel(
  serosurvey = serosurvey_large_epi,
  model_type = "time",
  is_log_foi = TRUE,
  foi_index = foi_index,
  iter = 2000
)

plot_log_time_normal <- plot_seromodel(
  seromodel_log_time_normal,
  serosurvey = serosurvey_large_epi,
  foi_df = foi_df,
  size_text = 5,
  foi_max = 0.7
)
plot(plot_log_time_normal)
``` 

In Fig 3 we can see that the *fast epidemic serofoi model* is able to identify the large epidemic simulated on the `simdata_large_epi` dataset.

## Models Comparison

Above we showed that the fast epidemic model (specified by `model_type = "time"` and `is_log = TRUE` in `fit_seromodel()`) is able to identify the large epidemic outbreak described by the dataset simulated according to a step-wise decreasing FoI (red line in Fig 3). 

Now, we would like to know whether this model actually fits this dataset better than the other available models in ***serofoi***. For this, we also implement both the endemic model (`model_type = "constant"`)   and the slow time-varying normal model (`model_type="time"`, `is_log = FALSE`):

```{r model_comparison, include = FALSE, echo = TRUE, results="hide", errors = FALSE, warning = FALSE, message = FALSE }
seromodel_constant <- fit_seromodel(
  serosurvey = serosurvey_large_epi,
  model_type = "constant",
  iter = 800
)
plot_constant <- plot_seromodel(
  seromodel_constant,
  serosurvey = serosurvey_large_epi,
  foi_df = foi_df,
  size_text = 5,
  plot_constant = TRUE,
  x_axis = "time",
  foi_max = 0.7
)

foi_index <- data.frame(
  year = seq(2000, 2049),
  foi_index = rep(c(1, 2, 3), c(30, 3, 17))
)
seromodel_time_normal <- fit_seromodel(
  serosurvey = serosurvey_large_epi,
  model_type = "time",
  foi_index = foi_index,
  iter = 2000
)
plot_time_normal <- plot_seromodel(
  seromodel_time_normal,
  serosurvey = serosurvey_large_epi,
  foi_df = foi_df,
  size_text = 5,
  foi_max = 0.7
)
```
Using the function `cowplot::plot_grid` we can visualise the results of the three models simultaneously:

```{r model_comparison_plot, fig.cap = "Figure 4. Model comparison between the three serofoi models for a large-epidemic simulated dataset.", include = TRUE, echo = TRUE, errors = FALSE, warning = FALSE, message = FALSE, fig.width=5, fig.asp=1, fig.align="center"}
cowplot::plot_grid(
  plot_constant, plot_time_normal, plot_log_time_normal,
  nrow = 1, ncol = 3, labels = "AUTO"
)
```

A common criterion to decide what model fits the data the best is to choose the one with the larger `elpd`. According to this criterion, in this case the epidemic and fast-epidemic models perform similarly (see the second row of panels B and C in Figure 4).

NOTE: Running the ***serofoi*** models for the first time on your local computer may take a few minutes for the rstan code to compile locally. However, once the initial compilation is complete, there is no further need for local compilation.

## References
